Generalizing the Model View Controller Paradigm

### 2006-11-21 inst/paper/MVC.tex
Made changes to paper 2.  Still needs some cutting.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@21240 bc3139a8-67e5-0310-9ffc-ced21a209358

@@

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset.  In some cases the
natural data representation of the subset is in a form other than that
of the initial data set.  This leads to complications in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a single, standard format when implementing the design.  We found that 
adding new views under such conditions could be quite problematic and 
often led to very complex code.  As we shall demonstrate, there are 
also issues that arise due to mappings that are not one-to-one between 
elements on different views.  Hence we considered an approach that 
created new instances of the MVC that were linked to the parent that 
spawned them.

In particular, our implementation consists of a hierarchy of MVCs,
stored as a tree.  Each parent can have multiple children, but the
children have a single parent.  Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow.  Thus, we can devise a number of purpose built MVC
classes, each with a model appropriate to the specific data of
interest, that can be linked in a natural way.  Interacting with the view
in a child MVC, first sends a message to the child MVC, which will
update all of its views, but the message will also propagate to the
parent MVC (and further up to the root).  Similarly, messages propagate 
to the child MVCs if the user initially interacts with a view in the 
parent MVC.  The MVC can choose to react, or not, to messages from its 
parent or children -- all control being in
the hands of the users, should they want it.

@@

%% FIXME: perhaps we could have a heatmap and a few of the other views
%% here?
%% EAW: put them below
There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (though edges could also be considered the primary 
unit), while
for the heatmap view the basic element is generally the gene.  In Figures
\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
graph are shown, respectively.  By interacting with the heatmap to color one
gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
GO nodes to be colored blue, one can easily see how this gene relates to the
GO terms.  Also by interacting with the node with an arrow next to it in
Figure \ref{Fig:GOgraph}, which represents the signal transducer activity GO
term, one can create a heatmap view that shows only the genes annotated at
that node, as shown in Figure \ref{Fig:heatmapsub}.  Notice how all genes
annotated at the signal transducer activity GO term have a blue label in
Figure \ref{Fig:heatmapsub}, which shows that this view was derived from a
blue colored node in Figure \ref{Fig:GOgraph}.  Compare the heatmaps in
Figures \ref{Fig:heatmapsub} and \ref{Fig:heatmap}.  The heatmap in Figure
\ref{Fig:heatmapsub} is just a subset of the heatmap in Figure
\ref{Fig:heatmap}, however in Figure \ref{Fig:heatmapsub} eleven genes have a
blue label and in Figure \ref{Fig:heatmap} only one gene has a blue label.
This difference in these heatmaps indicates that views are synchronized based
on the parent-child relationships.  Figure \ref{Fig:heatmap} is the parent of
Figure \ref{Fig:GOgraph}, which is the parent of Figure \ref{Fig:heatmapsub}.
Figures \ref{Fig:heatmap} and \ref{Fig:GOgraph} are synchronized, and Figures
\ref{Fig:GOgraph} and \ref{Fig:heatmapsub} are synchronized, but Figures
\ref{Fig:heatmap} and \ref{Fig:heatmapsub} are not synchronized by color
because they do not have a parent-child relationship: Figure \ref{Fig:heatmap}
is the grandparent of Figure \ref{Fig:heatmapsub}.

@@

\section{Methods}

We first give our definitions of linked views, interactive views and
linked data sets.  Our model is typically a statistical data
matrix which represents data on a number of variables, for a
number of samples.  By \textit{linked view} we mean that if a
component that represents data (such as a point on a plot or a row in
a spreadsheet) changes its appearance on one view, then the
corresponding component in all other views also changes appearance.
In our example, correspondence is determined by the sample,
but there are other possibilities, as we shall explain later.  As an
example, suppose that the model consists of the gene expression levels
from say 50 samples and 10,000 genes.  The model can be represented as
a matrix of 50 columns and ten thousand rows.  We have two choices for
linking, either the entities can be the samples, or they can be the
genes. If the entities are the genes then when two scatter plots are
created as views, where the first plot displays sample one versus sample two 
and the second displays sample three versus sample four, then the points
on each plot that referred to the same gene are corresponding
components.  If instead we had chosen to link on samples, then we
could produce pairwise plots of gene expression values and would link
them through the sample IDs.

%%FIXME: need to reference iPlots from Simon, and some of these refs need
%%to be put up at the front where I have a citations in parens
%%EAW: added a reference to iPlots below
Linked views, which have a long history in statistics, are now
considered a standard feature of interactive data visualization
software \cite{GGobi}.  Linked views have been implemented in several
pieces of statistical software, such as DataDesk \cite{DataDesk},
XLisp-Stat \cite{Lisp}, iPlots \cite{iPlots}, and
GGobi \cite{GGobiMan}.  Implementing linked views is based on the
model-view-controller (MVC) design \cite{DesignPatterns}, which is a
widely used and well understood paradigm.

We take a fairly general view of interactivity.  Users can interact via
a pointing device, such as a mouse, or via key strokes, or any of many
other types of input devices.  Interactivity means that there is an
immediate response to some user action.  It is a powerful paradigm and allows
users to quickly navigate and explore complex documents or complex
data sets.  As an example, suppose that a user has two views: one view
is a graph where each node represents a particular set of genes while
the other view is a heatmap of gene expression data.  If the user
clicks on a node then the response to this action could be to
highlight all genes in the heatmap view that correspond to the
selected node.

@@

Consider Figure \ref{Fig:firstMP}, which provides a cartoon
description of the steps that occur when a user interacts with a
view.  In the cartoon there are only two views, one of model A (note 
that once a data set is loaded in a MVC, we call it a model) and the
other of the linked model B.  When a user
interacts with the scatter plot view of model B, the response to
this interaction is for the controller to tell model B to update itself 
(this is indicated
on the figure as a circled 1).  In Step 2, the scatter plot view is
told to update itself.  Next, in Step 3, because model B is linked
to model A, model B sends a message to model A.  In Step 4,
the information from model B is converted into information that
model A understands.  In Step 5, this information is received by
model A and model A updates itself.  Finally, in Step 6, model
A tells its view to update itself.

% In the design presented in this paper, the components consist of
% \Robject{MVC} objects and communication within and between
% \Robject{MVC} objects is performed by message objects.  This solution
% uses the power and simplicity of the MVC design and expands it to the
% multiple data set problem.  Now we can build on the original MVC
% design, which has been well tested, instead of starting a new design
% from scratch.  The strength of this multiple linked MVC design is that
% it uses MVC components, which can be individually created and then
% linked.  Thus, these components are reusable.  These design decisions
% are discussed further in Sections \ref{Sec:OneMVC} and
% \ref{Sec:MultMVC}.

@@

% In Section \ref{Ssec:Design}, the design of linked, interactive views based on
% the MVC paradigm is discussed.  However, the limitation with this design is
% that it expects one data set, which limits the type of data we can visualize.
% There are many situations where multiple data sets occur, particularly any
% time there are experimental or study data linked to meta-data.  Being able to
% visualize data from multiple data sets is helpful to understand the
% relationships between variables.  Thus, after discussing the limitations of
% the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
% discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
% linked \Robject{MVC} objects.  The design of individual \Robject{MVC} objects
% is discussed in 
% Section \ref{Sec:OneMVC} while the implementation of linked \Robject{MVC}
% objects is discussed in Section \ref{Sec:MultMVC}.  The goals of the software
% discussed in this paper are to create linked views of linked data sets, to
% create interactive views where the response to an event can be changed, and to
% create an extensible design that allows future users to make additions.

@@

\subsection{Limitations of the MVC Design}\label{Ssec:Limit}

In essence the MVC design assumes that there is one data set, which is
called the model, that all views reference.  But problems arise when
the data consist of more than one conceptual grouping.  For example,
consider data stored in a relational database, such as the data shown in
Figure~\ref{Fig:DBTab}.  Here the database
consists of three tables, State, City and StateCity.  The State table
has three columns, State (the key), Life Expectancy, and Income.  The
City table has three columns, City (the key), Precipitation, and
Temperature.  The StateCity table has two columns, State and City.
This last table relates which cities are in which states.  In this
example, each city can be in only one state, but each state can have
more than one city so there is a one-to-many relationship between
states and cities.  In order to support interactive views of all data
tables, in a standard MVC paradigm, it seems that either all data must
be combined into a single large table (Figure~\ref{Fig:OneDBTab}), or
the processing and accessing of the model data must become much more
complex.

@@

\begin{figure}[ht]
  \begin{center}
    \subfigure[Example of relational database
    tables.]{
    \label{Fig:DBTab}
    \includegraphics[height=1.2in, width=3.5in]{newDatabaseTables.jpg}}
    \hspace{.3in}
    \subfigure[Example of a combined table.]{
    \label{Fig:OneDBTab}
    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}}
    \caption{Relational database tables.}
%    \label{Fig:DBTab}
  \end{center}
\end{figure}

We felt that neither approach had much appeal.  A single large table
suggests that all data are similar, and still leaves issues about how
to deal with ensuring that when, for example, a state is selected all
cities within it are highlighted.  The second approach requires writing a
large amount of code, but is unlikely to be extensible.
We should also mention that our main area of
application is a setting where one representation of the data is as a
graph (with nodes and edges) while the other is as a rectangular array. 
We can
envisage lots of different MVCs that operate on graph based data, and
equally MVCs that operate on matrix based data.  Specifying a mechanism
by which these can interact allows us to create two visible classes
that others can use, but still provide interaction between
appropriately linked instances of the two classes.  We also argue that
there are surely other specific types of models that represent
specific types of data (e.g. time series) and it would make sense to
implement these as yet another separate, specialized MVC, that could
be linked to any one of a variety of other MVCs.  In our design, each 
MVC will contain one model and thus, linked models also indicate linked 
MVCs.

@@

\section{Extending the MVC Design}
\label{Sec:Extend}

As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
conceptual unit, such as a data frame or a matrix, for the model.  And
we believe that such an approach is a good one.  MVCs should be
specialized to certain types of data, as this allows them to provide
better and more natural interactions.  However, we additionally argue
that there is great benefit to considering linking these different
MVCs when the underlying data themselves are derived from the same source.
We believe that there will often be a hierarchical relationship
between the models and the MVCs that contain them.  We have found
that allowing the user to create new MVCs based on a current MVC is an
important and useful paradigm.

Each \Robject{MVC} is a coherent object and the interactions between
\Robject{MVCs} are encapsulated in a set of messages.  We use one
\Robject{MVC} object to handle each model.  Thus, within a
\Robject{MVC} object, one can create linked views of that particular
model, using the standard MVC design.  Between \Robject{MVC}
objects we propose a second messaging system that allows messages to
be sent between linked models.

@@

@@

In Figure \ref{Fig:MultMVC} we show one view for state-level data and
one view for the city-level data.  The state and city data are linked
based on the color.  But there are two distinct MVCs, not one. 
The city data have one MVC and the state data a second.  Even though 
there are two \Robject{MVC} objects here, the views are linked.

\section{MVC Design Details}
\label{Sec:OneMVC}

\subsection{Implementation Details}
\label{SSec:OneOver}

We have implemented linked, interactive views of linked data sets
using an object-oriented programming design which allows for
extensibility.  We provide this implementation in the
\Rpackage{MVCClass} package, written in R, which implements the basic
classes and generic functions.  The \Rpackage{MVCClass} package
defines the individual model, view, and controller classes as well as
the \Rclass{MVC} class that binds the individual model, view, and
controller classes together into one unit.  The package also defines
the message classes that allow the components to communicate with each
other.  Thus, the \Rpackage{MVCClass} package contains the classes
necessary for creating linked views.

@@

% Figure \ref{Fig:Model} shows the inheritance structure for the model
% classes. Here the virtual class, \Rclass{gModel}, has the slots
% \Rslot{modelData}, \Rslot{linkData}, \Rslot{virtualData}, and
% \Rslot{modelName}.  These four slots are common information that all models
% need to store.  The \Rslot{modelData} slot is the data set for this model, the
% \Rslot{linkData} slot is a list of two functions, \Rfunction{toParent} and
% \Rfunction{fromParent}, which let this model send information to its parent
% model and receive information from its parent model, respectively (see
% Sections \ref{Ssec:OneMVC} and \ref{Ssec:MultLink} for more information on how
% parent and child models are linked), the \Rslot{virtualData} slot is data
% pertaining to the views that needs to be stored with the model so that it can
% be shown in all views of this model, and the \Rslot{modelName} slot is the
% name of the model.

@@

Although only three model classes are currently implemented, future
users are welcome to make additions to the model classes for new data
structures they want to represent.  For example, a user may want to add a new
model class called \Rclass{tsModel} to represent time series data.
To make this addition, the user extends the \Rclass{gModel} class in his or
her own R software package.  More information on extending the classes can be
found in the \Rpackage{MVCClass} and \Rpackage{iSNetwork} Vignettes.

@@

The controller performs two tasks: the first task is to maintain state,
such as what response will occur for a particular user interaction, 
and the second task is to control message passing.  Messages, which are
discussed in Sections \ref{Ssec:OneMess} and \ref{Ssec:MultLink}, are
responsible for creating new components, such as models and views, and for
passing information between components, such as a view asking a model for
information.  The controller determines when these messages are created and
ensures that they are sent to the proper recipient.

@@

\subsection{MVC}
\label{Ssec:OneMVC}

The \Rclass{MVC} class binds the model, view, and controller objects
into one object.  Thus, whenever a new data set is loaded, an
appropriate \Robject{MVC} object is created.  There is a one-to-one
relationship between the \Robject{MVC} instances and model objects.
%%FIXME: this seems like an implementation detail
%%Because of this relationship, the \Robject{MVC} object and the model
%%object are identified by the same name.
Our implementation consists
of a virtual class, \Rclass{MVC}, with three slots, which are
\Rslot{model}, \Rslot{viewList}, and \Rslot{controller}. The three
slots correspond to the three pieces of information that are needed to
bind the model, view, and controller into a single unit. This class is
sufficient for all single MVC activities. To support linked models
we extend this class with the \Rclass{linkedModelMVC} class, which has
two additional slots: \Rslot{parentMVC}, which is the name of the
parent model and \Rslot{childMVCList}, which is a list of the child models.
These are the models that will communicate with each other: parents send
information to their children and children send information to their parent.
More information on this communication is given in Section \ref{Sec:MultMVC}.

@@

As shown in Figure \ref{Fig:Hier}, there are many cases where there are
natural submodels that can be derived from a parent model.  We propose that
each submodel is itself a \Robject{MVC} instance, and hence can have multiple
linked views.  Even though each submodel is its own \Robject{MVC} instance,
there will be times when the submodel \Robject{MVC} needs to communicate with
its parent \Robject{MVC}.  There are many different ways of communicating
between parent and child \Robject{MVCs} and we discuss this in Section
\ref{Sec:MultMVC}.

>>>>

\setlength{\baselineskip}{20pt}

@@

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset.  In some cases the
natural data representation of the subset is in a form other than that
of the initial data set.  This leads to complications in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a single, standard format when implementing the design.  We found that 
adding new views under such conditions could be quite problematic and 
often led to very complex code.  As we shall demonstrate, there are 
also issues that arise due to mappings that are not one-to-one between 
elements on different views.  Hence we considered an approach that 
created new MVC instances that were linked to the parent MVC that 
spawned them.

In particular, our implementation consists of a hierarchy of MVCs,
stored as a tree, as shown in Figure \ref{Fig:Hier}.  Each parent can 
have multiple children, but the
children have a single parent.  Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow.  Thus, we can devise a number of purpose built MVC
classes, each with a model appropriate to the specific data of
interest, that can be linked in a natural way.  Interacting with the view
in a child MVC, first sends a message to the child MVC, which will
update all of its views, but the message will also propagate to the
parent MVC (and further up to the root).  Similarly, messages propagate 
to the child MVCs if the user initially interacts with a view in the 
parent MVC.  The MVC can choose to react, or not, to messages from its 
parent or children -- all control being in
the hands of the users, should they want it.

@@

%% FIXME: perhaps we could have a heatmap and a few of the other views
%% here?
%% EAW: put them below
There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (though edges could also be considered the primary 
unit), while
for the heatmap view the basic element is generally the gene.  In Figures
\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
graph are shown, respectively.  By interacting with the heatmap to color one
gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
GO nodes to be colored blue, one can easily see how this gene relates to the
GO terms.  Also by interacting with the node with an arrow next to it in
Figure \ref{Fig:GOgraph}, which represents the signal transducer activity GO
term, one can create a heatmap view that shows only the genes annotated at
that node, as shown in Figure \ref{Fig:heatmapsub}.  Notice how all genes
annotated at the signal transducer activity GO term have a blue label in
Figure \ref{Fig:heatmapsub}, which shows that this view was derived from a
blue colored node in Figure \ref{Fig:GOgraph}.  Compare the heatmaps in
Figures \ref{Fig:heatmapsub} and \ref{Fig:heatmap}.  The heatmap in Figure
\ref{Fig:heatmapsub} is just a subset of the heatmap in Figure
\ref{Fig:heatmap}, however in Figure \ref{Fig:heatmapsub} eleven genes have a
blue label and in Figure \ref{Fig:heatmap} only one gene has a blue label.
The difference in these heatmaps indicates that views are synchronized based
on the parent-child relationships.  Figure \ref{Fig:heatmap} is the parent of
Figure \ref{Fig:GOgraph}, which is the parent of Figure \ref{Fig:heatmapsub}.
Figures \ref{Fig:heatmap} and \ref{Fig:GOgraph} are synchronized, and Figures
\ref{Fig:GOgraph} and \ref{Fig:heatmapsub} are synchronized, but Figures
\ref{Fig:heatmap} and \ref{Fig:heatmapsub} are not synchronized by color
because they do not have a parent-child relationship: Figure \ref{Fig:heatmap}
is the grandparent of Figure \ref{Fig:heatmapsub}.

@@

In this paper we describe an extensible design for creating linked,
interactive views of linked data sets using an extension of the MVC 
paradigm.  Section \ref{Sec:Back} gives definitions of the terms used in this
paper and describes the MVC paradigm along with its
limitations.  Then Section \ref{Sec:OneMVC} describes our implementation of
the MVC paradigm and how it was extended to accommodate linked data sets.

\section{Background}
\label{Sec:Back}

\subsection{Definitions}
\label{Ssec:Def}

We first give our definitions of linked views, and
linked data sets.  Our model is typically a statistical data
matrix which represents data on a number of variables, for a
number of samples.  By \textit{linked view} we mean that if a
component that represents data (such as a point on a plot or a row in
a spreadsheet) changes its appearance on one view, then the
corresponding component in all other views also changes appearance.
In our example, correspondence is determined by the sample,
but there are other possibilities, as we shall explain later.  As an
example, suppose that the model consists of the gene expression levels
from say 50 samples and 10,000 genes.  The model can be represented as
a matrix of 50 columns and ten thousand rows.  We have two choices for
linking, either the entities can be the samples, or they can be the
genes. If the entities are the genes then when two scatter plots are
created as views, where the first plot displays sample one versus sample two 
and the second displays sample three versus sample four, then the points
on each plot that referred to the same gene are corresponding
components.  If instead we had chosen to link on samples, then we
could produce pairwise plots of gene expression values and would link
them through the sample IDs.

%%FIXME: need to reference iPlots from Simon, and some of these refs need
%%to be put up at the front where I have a citations in parens
%%EAW: added a reference to iPlots below
Linked views, which have a long history in statistics, are now
considered a standard feature of interactive data visualization
software \cite{GGobi}.  Linked views have been implemented in several
pieces of statistical software, such as DataDesk \cite{DataDesk},
XLisp-Stat \cite{Lisp}, iPlots \cite{iPlots}, and GGobi \cite{GGobiMan}.  
Implementing linked views is generally based on the MVC paradigm
\cite{DesignPatterns}, which is widely used and well understood.  We describe
it in detail in Section \ref{Ssec:MVC}.

%We take a fairly general view of interactivity.  Users can interact via
%a pointing device, such as a mouse, or via key strokes, or any of many
%other types of input devices.  Interactivity means that there is an
%immediate response to some user action.  It is a powerful paradigm and allows
%users to quickly navigate and explore complex documents or complex
%data sets.  As an example, suppose that a user has two views: one view
%is a graph where each node represents a particular set of genes while
%the other view is a heatmap of gene expression data.  If the user
%clicks on a node then the response to this action could be to
%highlight all genes in the heatmap view that correspond to the
%selected node.

@@

\subsection{Model View Controller (MVC) Paradigm}
\label{Ssec:MVC}

The MVC paradigm \citep{DesignPatterns} is a design that consists of
three types of objects: the controller, which defines what actions
occur in response to user input; the view, which consists of displays
of the data; and the model, which manages the data.  The power of the
MVC design is that it decouples the views of the model from the model
itself by creating a subscribe/notify procedure between them. In other
words, the views subscribe to a particular model and the model must
notify the views when a change occurs so that the views can be
updated.  This allows multiple views of the model that are linked, but
are not aware of each other.  And similarly, the model has no
particular knowledge of any of the views, just how to notify them of
changes to the contents of the model.

In the MVC paradigm the components are separated, but they need a 
method to communicate with each other, which can be implemented using 
a messaging system.  If we have linked data sets, then Figure
\ref{Fig:firstMP} displays the steps that are needed to synchronize 
each model with its views as well as synchronizing the linked data sets 
when a user interacts with a view.
In the figure there are only two views, one of model A (note 
that once a data set is loaded in a MVC, we call it a model) and the
other of the linked model B.  When a user
interacts with the scatter plot view of model B, the response to
this interaction is for the controller to tell model B to update itself 
(this is indicated
on the figure as a circled 1).  In Step 2, the scatter plot view is
told to update itself.  Next, in Step 3, because model B is linked
to model A, model B sends a message to model A.  In Step 4,
the information from model B is converted into information that
model A understands.  In Step 5, this information is received by
model A and model A updates itself.  Finally, in Step 6, model
A tells its view to update itself.

% In the design presented in this paper, the components consist of
% \Robject{MVC} objects and communication within and between
% \Robject{MVC} objects is performed by message objects.  This solution
% uses the power and simplicity of the MVC design and expands it to the
% multiple data set problem.  Now we can build on the original MVC
% design, which has been well tested, instead of starting a new design
% from scratch.  The strength of this multiple linked MVC design is that
% it uses MVC components, which can be individually created and then
% linked.  Thus, these components are reusable.  These design decisions
% are discussed further in Sections \ref{Sec:OneMVC} and
% \ref{Ssec:MultMVC}.

@@

% In Section \ref{Ssec:Design}, the design of linked, interactive views based on
% the MVC paradigm is discussed.  However, the limitation with this design is
% that it expects one data set, which limits the type of data we can visualize.
% There are many situations where multiple data sets occur, particularly any
% time there are experimental or study data linked to meta-data.  Being able to
% visualize data from multiple data sets is helpful to understand the
% relationships between variables.  Thus, after discussing the limitations of
% the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
% discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
% linked \Robject{MVC} objects.  The design of individual \Robject{MVC} objects
% is discussed in 
% Section \ref{Sec:OneMVC} while the implementation of linked \Robject{MVC}
% objects is discussed in Section \ref{Ssec:MultMVC}.  The goals of the software
% discussed in this paper are to create linked views of linked data sets, to
% create interactive views where the response to an event can be changed, and to
% create an extensible design that allows future users to make additions.

@@

\subsection{Limitations of the MVC Paradigm}\label{Ssec:Limit}

When implementing the MVC paradigm, it is often assumed that there is 
one data set, which is called the model, that all views reference.  
But problems arise when the data consist of more than one conceptual 
grouping.  For example, consider data stored in a relational database, 
such as the data shown in Figure \ref{Fig:MultMVC}.
%Figure~\ref{Fig:DBTab}.  
Here the database
consists of three tables, State, City and StateCity.  The State table
has three columns, State (the key), Life Expectancy, and Income.  The
City table has three columns, City (the key), Precipitation, and
Temperature.  The StateCity table has two columns, State and City.
This last table relates which cities are in which states.  In this
example, each city can be in only one state, but each state can have
more than one city so there is a one-to-many relationship between
states and cities.  In order to support interactive views of all data
tables, in a standard MVC implementation, it seems that either all 
data must be combined into a single large table, or
the processing and accessing of the model data must become much more
complex.  Notice that the MVC paradigm allows a model that is complex, but
accessing the model data is more straightforward when the data is a single
conceptual unit.  This single data structure as a model creates a well 
defined message passing system between the model and its views as well as
between the model and the controller.

@@

%\begin{figure}[ht]
%  \begin{center}
%    \subfigure[Example of relational database
%    tables.]{
%    \label{Fig:DBTab}
%    \includegraphics[height=1.2in, width=3.5in]{newDatabaseTables.jpg}}
%    \hspace{.3in}
%    \subfigure[Example of a combined table.]{
%    \label{Fig:OneDBTab}
%    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}}
%    \caption{Relational database tables.}
%%    \label{Fig:DBTab}
%  \end{center}
%\end{figure}

We felt that neither approach of combining linked data into one data 
structure or of having a complex model had much appeal.  A single large 
table suggests that all data are similar, and still leaves issues about how
to deal with ensuring that when, for example, a state is selected all
cities within it are highlighted.  The second approach requires writing a
large amount of code, but is unlikely to be extensible.  One area of 
interest is a setting where one representation of the data is as a
graph (with nodes and edges) while the other is as a rectangular array. 
We can envisage lots of different MVCs that operate on graph based data, and
equally MVCs that operate on matrix based data.  Specifying a mechanism
by which these can interact allows us to create two visible classes
that others can use, but still provide interaction between
appropriately linked instances of the two classes.  We also argue that
there are other specific types of models that represent
specific types of data (e.g. time series) and it would make sense to
implement these as other specialized MVCs.

@@

%\subsection{Extending the MVC Design}
%\label{Ssec:Extend}

%As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a 
%single
%conceptual unit, such as a data frame or a matrix, for the model.  And
%we believe that such an approach is a good one.  
%We believe that MVCs should be specialized to certain types of data, such 
%as a graph or a data frame, as this allows them to provide
%better and more natural interactions.  However, we additionally argue
%that there is great benefit to considering linking these different
%MVCs when the underlying data themselves are derived from the same source.
%We believe that there will often be a hierarchical relationship
%between the models and the MVCs that contain them.  We have found
%that allowing the user to create new MVCs based on a current MVC is an
%important and useful paradigm.

%Each \Robject{MVC} is a coherent object and the interactions between
%\Robject{MVCs} are encapsulated in a set of messages.  We use one
%\Robject{MVC} object to handle each model.  Thus, within a
%\Robject{MVC} object, one can create linked views of that particular
%model, using the standard MVC design.  Between \Robject{MVC}
%objects we propose a second messaging system that allows messages to
%be sent between linked models.  In Figure \ref{Fig:MultMVC} we show one 
%view for state-level data and one view for the city-level data.  The state 
%and city data are linked based on the color.  But there are two distinct 
%MVCs, not one.  The city data have one MVC and the state data a second.  
%Even though there are two \Robject{MVC} objects here, the views are linked.

@@

%\begin{figure}[ht]
%  \begin{center}
%    \includegraphics[height=5in, width=6in]{MultipleMVC.jpg}
%    \caption{ Example of two \Robject{MVC} objects.  These two \Robject{MVC}
%      objects are linked by the StateCity table and the color of the points
%    indicates the linking between the data.  The points referring to the state
%    of Pennsylvania are red, the points referring to the state of California
%    are blue, and the points referring to Minnesota are green. }
%    \label{Fig:MultMVC}
%  \end{center}
%\end{figure}

%In Figure \ref{Fig:MultMVC} we show one view for state-level data and
%one view for the city-level data.  The state and city data are linked
%based on the color.  But there are two distinct MVCs, not one. 
%The city data have one MVC and the state data a second.  Even though 
%there are two \Robject{MVC} objects here, the views are linked.

\section{MVC Implementation Details}
\label{Sec:OneMVC}

We believe that MVCs should be specialized to certain types of data, such 
as a graph or a data frame, as this allows them to provide
better and more natural interactions.  However, we additionally argue
that there is great benefit to considering linking these different
MVCs when the underlying data themselves are derived from the same source.
We believe that there will often be a hierarchical relationship
between the models and the MVCs that contain them.  We have found
that allowing the user to create new MVCs based on a current MVC is an
important and useful paradigm.

Each \Robject{MVC} is a coherent object and the interactions between
\Robject{MVCs} are encapsulated in a set of messages.  We use one
\Robject{MVC} object to handle each model.  Thus, within a
\Robject{MVC} object, one can create linked views of that particular
model, using the standard MVC design.  Between \Robject{MVC}
objects we propose a second messaging system that allows messages to
be sent between linked models.  In Figure \ref{Fig:MultMVC} we show one 
view for state-level data and one view for the city-level data.  The state 
and city data are linked based on the color.  But there are two distinct 
MVCs, not one.  The city data have one MVC and the state data a second.  
Even though there are two \Robject{MVC} objects here, the views are linked.

@@

%\subsection{Implementation Details}
%\label{SSec:OneOver}
%
%We have implemented linked, interactive views of linked data sets
%using an object-oriented programming design which allows for
%extensibility.  We provide this implementation in the
%\Rpackage{MVCClass} package, written in R, which implements the basic
%classes and generic functions.  The \Rpackage{MVCClass} package
%defines the individual model, view, and controller classes as well as
%the \Rclass{MVC} class that binds the individual model, view, and
%controller classes together into one unit.  The package also defines
%the message classes that allow the components to communicate with each
%other.  Thus, the \Rpackage{MVCClass} package contains the classes
%necessary for creating linked views.

@@

% Figure \ref{Fig:Model} shows the inheritance structure for the model
% classes. Here the virtual class, \Rclass{gModel}, has the slots
% \Rslot{modelData}, \Rslot{linkData}, \Rslot{virtualData}, and
% \Rslot{modelName}.  These four slots are common information that all models
% need to store.  The \Rslot{modelData} slot is the data set for this model, the
% \Rslot{linkData} slot is a list of two functions, \Rfunction{toParent} and
% \Rfunction{fromParent}, which let this model send information to its parent
% model and receive information from its parent model, respectively (see
% Sections \ref{Ssec:OneMVC} and \ref{Sssec:MultLink} for more information on how
% parent and child models are linked), the \Rslot{virtualData} slot is data
% pertaining to the views that needs to be stored with the model so that it can
% be shown in all views of this model, and the \Rslot{modelName} slot is the
% name of the model.

@@

Although only three model classes are currently implemented, future
users are welcome to make additions to the model classes for new data
structures they want to represent.  For example, a user may want to add a new
model class called \Rclass{tsModel} to represent time series data.
To make this addition, the user extends the \Rclass{gModel} class, which is
the general model class, in his or
her own R software package.  More information on extending the classes can be
found in the \Rpackage{MVCClass} and \Rpackage{iSNetwork} vignettes, which are
supplied with the software packages.

@@

The controller performs two tasks: the first task is to maintain state,
such as what response will occur for a particular user interaction, 
and the second task is to control message passing.  Messages, which are
discussed in Sections \ref{Ssec:OneMess} and \ref{Sssec:MultLink}, are
responsible for creating new components, such as models and views, and for
passing information between components, such as a view asking a model for
information.  The controller determines when these messages are created and
ensures that they are sent to the proper recipient.

@@

\subsection{Putting it all Together in a MVC Class}
\label{Ssec:OneMVC}

The \Rclass{MVC} class binds the model, view, and controller objects
into one object.  Thus, whenever a new data set is loaded, an
appropriate \Robject{MVC} object is created.  There is a one-to-one
relationship between the \Robject{MVC} instances and model objects.
%%FIXME: this seems like an implementation detail
%%Because of this relationship, the \Robject{MVC} object and the model
%%object are identified by the same name.
Our implementation consists
of a virtual class, \Rclass{MVC}, with three slots, which are
\Rslot{model}, \Rslot{viewList}, and \Rslot{controller}. The three
slots correspond to the three pieces of information that are needed to
bind the model, view, and controller into a single unit. This class is
sufficient for all single MVC activities. To support linked models
we extend this class with the \Rclass{linkedModelMVC} class, which has
two additional slots: \Rslot{parentMVC}, which is the name of the
parent model and \Rslot{childMVCList}, which is a list of the child models.
These are the models that will communicate with each other: parents send
information to their children and children send information to their parent.
More information on this communication is given in Section \ref{Ssec:MultMVC}.

@@

%\begin{figure}[ht]
%  \begin{center}
%    \includegraphics[height=2.5in, width=4.2in]{Hierarchy2.jpg}
%    \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
%    \label{Fig:Hier}
%  \end{center}
%\end{figure}

As shown in Figure \ref{Fig:Hier}, there are many cases where there are
natural submodels that can be derived from a parent model.  We propose that
each submodel is itself a \Robject{MVC} instance, and hence can have multiple
linked views.  Even though each submodel is its own \Robject{MVC} instance,
there will be times when the submodel \Robject{MVC} needs to communicate with
its parent \Robject{MVC}.  There are many different ways of communicating
between parent and child \Robject{MVCs} and we discuss this in Section
\ref{Ssec:MultMVC}.
